---
title: "p8105_hw3_iho2104"
author: "Ixtaccihuatl Obregon"
date: "`r Sys.Date()`"
output: github_document
---


```{r}
library(tidyverse)
library(ggridges)
library(patchwork)
library(dplyr)
library(p8105.datasets)
```

## Problem 1 

```{r}
data("instacart")
instacart = 
  instacart |> 
  as_tibble()
```

Description: There are 1384617 observations and 15 columns/variables. Important variables include product_id, user_id, aisle_id and product_name. 

```{r}
instacart |> 
  count(aisle) |> 
  arrange(desc(n))
```

How many aisles are there, and which aisles are the most items ordered from? 
There are 134 aisles and most items are fresh fruits and vegetables. 


Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.

```{r}
instacart |> 
  count(aisle) |> 
  filter(n > 10000) |> 
  mutate(aisle = fct_reorder(aisle, n)) |> 
  ggplot(aes(x = aisle, y = n)) + 
  geom_point() + 
  labs(title = "Number of items ordered in each aisle") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.

```{r}
instacart |> 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |>
  group_by(aisle) |> 
  count(product_name) |> 
  mutate(rank = min_rank(desc(n))) |> 
  filter(rank < 4) |> 
  arrange(desc(n)) |>
  knitr::kable()
```

Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).

```{r}
instacart |>
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |>
  group_by(product_name, order_dow) |>
  summarize(mean_hour = mean(order_hour_of_day)) |>
  pivot_wider(
    names_from = order_dow, 
    values_from = mean_hour) |>
  knitr::kable(digits = 2)
```


## Problem 2 

```{r}
data("brfss_smart2010")
```

First, do some data cleaning:
- format the data to use appropriate variable names;
- focus on the “Overall Health” topic
- include only responses from “Excellent” to “Poor”
- organize responses as a factor taking levels ordered from “Poor” to “Excellent”

```{r}
brfss_df =  
  brfss_smart2010 |> 
  janitor::clean_names() |> 
  filter(str_detect("Overall Health", topic)) |> 
  arrange(factor(response, levels = c("Poor", "Fair", "Very Good", "Good", "Excellent")))
```

 


```{r}
brfss_df |> 
  filter(year == 2002) |> 
  group_by(locationabbr) |> 
  summarize(observations =n()) |> 
  filter(observations >= 7) 
```
- In 2002, 36 states were observed at 7 or more locations.
36 states being AZ, CO, CT, DE, FL, GA, HI, ID, IL, IN, KS, LA, MA, MD, ME, MI, MN, MO, NC, NE, NH,NJ, NV, NY, OH, OK, OR, PA, RI, SC, SD, TN, TX, UT, VT, and WA.

```{r}
 brfss_df |> 
  filter(year == 2010) |> 
  group_by(locationabbr) |> 
  summarize(observations =n()) |> 
  filter(observations >= 7) 
```

In 2010, 45 locations consisting of: AL, AR, AZ, CA, CO, CT, DE, FL, GA, HI, IA, ID, IL, IN, KS, LA, MA, MD, ME, MI, MN, MO, MS, MT, NC, ND, NE, NH, NJ, NM, NV, NY, O, OK, OR, PA, RI, SC, SD, TN, TX, UT, VT, WA, and WY, had more than 7 or more locations .

- Construct a dataset that is limited to Excellent responses, and contains, year, state, and a variable that averages the data_value across locations within a state. 
- Make a “spaghetti” plot of this average value over time within a state (that is, make a plot showing a line for each state across years – the geom_line geometry and group aesthetic will help).

```{r}
brfss_excellent = 
  brfss_df |> 
  filter(str_detect("Excellent", response)) |> 
  group_by(locationabbr, year) |> 
  summarize(avg_data_value = mean(data_value, na.rm = TRUE) ) |> 
  select(year, locationabbr, avg_data_value ) |> 
  ggplot(aes(x = year, 
             y = avg_data_value, color = locationabbr))+
  geom_line() +
  labs(title = "average value over time within a state",
       x = "year",
       y = "mean") +
  viridis::scale_color_viridis(discrete = T )
brfss_excellent
```

- Make a two-panel plot showing, for the years 2006, and 2010, distribution of data_value for responses (“Poor” to “Excellent”) among locations in NY State.

```{r}
brfss_ny = brfss_df |> 
  filter(locationabbr == "NY", year %in% c(2006, 2010)) |> 
  ggplot( aes(
            x = factor(response, levels = c("Poor", "Fair", "Good", "Very good", "Excellent")),  
             y = data_value, 
            color = locationdesc)) +
  geom_point() +
  facet_wrap(~ year, ncol = 2) +
  labs(title = "Distribution of data_value by Response in NY State Counties",
       x = "Response",
       y = "Data Value") +
  viridis::scale_color_viridis(discrete = T )
brfss_ny
```

## Problem 3 

Load, tidy, merge, and otherwise organize the data sets. Your final dataset should include all originally observed variables; exclude participants less than 21 years of age, and those with missing demographic data; and encode data with reasonable variable classes (i.e. not numeric, and using factors with the ordering of tables and plots in mind).

```{r}
nhanes_demo = read_csv("Data/nhanes_covar.csv", skip = 3) |> 
  janitor::row_to_names(row_number = 1) |> 
  janitor::clean_names() |> 
   mutate(
    sex = 
      case_match(
        sex, 
        "1" ~ "male", 
        "2" ~ "female"),
    sex = as.factor(sex), 
    education = 
      case_match(
        education, 
        "1" ~ "less than high school", 
        "2" ~ "high school equivalent", 
        "3" ~ "more than high school"),
    education = as.factor(education), 
    seqn = as.numeric(seqn),
    age = as.numeric(age),
    bmi = as.numeric(bmi)) |> 
  filter(age >= 21) |> 
  na.omit(BMI, education) 

nhanes_accel = read_csv("Data/nhanes_accel.csv") |> 
  janitor::clean_names() |> 
  mutate(seqn = as.numeric(seqn)) 

nhanes_df = inner_join(nhanes_demo,nhanes_accel, by = "seqn")
```

Produce a reader-friendly table for the number of men and women in each education category, and create a visualization of the age distributions for men and women in each education category. Comment on these items.

```{r}
nhanes_df |> 
  count(sex, education)  |> 
  pivot_wider(
    names_from = education, 
    values_from = n)
```
```{r}
nhanes_df |> 
  ggplot(aes(
    x = factor(education, levels = c("more than high school", "high school equivalent", "less than high school")),
    y = age, 
    fill = sex
  )) +
  geom_boxplot()+ 
  labs(title = "age distributions for men and women in each education category",
       x = "education",
       y = "age") +
  viridis::scale_color_viridis(discrete = T )  
```

Comment: The mean age is similar between females and males in the "less than high school" and "more than high school" education groups. On average, females were older than men in the "high school equivalent group". 

Using your tidied dataset, aggregate across minutes to create a total activity variable for each participant. Plot these total activities (y-axis) against age (x-axis); your plot should compare men to women and have separate panels for each education level. Include a trend line or a smooth to illustrate differences. Comment on your plot.

```{r}
nhanes_df |> 
  mutate(total_activity = rowSums(across(min1:min1440))) |> 
  ggplot( aes(
            x = age,  
             y = total_activity, 
            color = sex)) +
  geom_point() +
  geom_smooth(method = lm) + 
  facet_wrap(~ factor(education, levels = c("more than high school", "high school equivalent", "less than high school")), ncol = 3) +
  labs(title = "Total activity vs age in men and women in different education levels",
       x = "age",
       y = "total activity") +
  viridis::scale_color_viridis(discrete = T )
```

Comment: Females and males with high school equivalent or more than high school education are seen to maintain there activity levels as they age. Those in the less than high school education show a decline in total activity as they age, with females showing the bigger decrease trend. 

Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex. Describe in words any patterns or conclusions you can make based on this graph; including smooth trends may help identify differences.

```{r}
nhanes_df |> 
  pivot_longer(
    min1:min1440, 
    names_to = "minute", 
    values_to = "activity") |> 
  mutate(minute = as.numeric(str_sub(minute,4, -1))) |> 
  ggplot(aes(
    x = minute, 
    y = activity,
    color = sex
  )) +
    geom_point(alpha = 1/10, size = 0.2)+
    geom_smooth()+
    facet_wrap(~ factor(education, levels = c("more than high school", "high school equivalent", "less than high school")), ncol = 3) +
  labs(title = "24-hour activity time courses for each education level",
       x = "minute of day",
       y = "activity") +
  viridis::scale_color_viridis(discrete = T )
```

Comment: Individuals showed a similar trend regardless of sex and/or education. However, in the "more than high school" education, the females showed more of a plateau trend starting at min500 than their male counterparts. The female individuals have a slightly higher activity level which could help idenitify some differences. 